High-level cognition is supported by at least second order dynamic correlations in neural activity patterns

### 2019-09-04 paper/main.tex
moved patterns figure to intro

\section*{Introduction}
A central goal in cognitive neuroscience is to elucidate the
\textit{neural code}: the mapping between (a) mental states or
cognitive representations and (b) neural activity patterns. One means
of testing models of the neural code is to ask how accurately that
model is able to ``translate'' neural activity patterns into known (or
hypothesized) mental states or cognitive
representations~\citep[e.g.,][]{HaxbEtal01, NormEtal06, TongPrat12,
  MitcEtal08a, KamiTong05, NishEtal11, PereEtal18, HuthEtal12,
  HuthEtal16}.  Training decoding models on different types of neural
features can also help to elucidate which specific aspects of neural
activity patterns are informative about cognition-- and, by extension,
which types of neural activity patterns might comprise the neural
code.  For example, prior work has used region of interest analyses to
estimate the anatomical locations of specific neural
representations~\citep[e.g.,][]{EtzeEtal09}, or to compare the
relative contributions to the neural code of multivariate activity
patterns versus patterns of dynamic correlations between neural
activity patterns~\citep[e.g.,][]{MannEtal18, FongEtal19}.  An
emerging theme in this literature is that cognition is mediated by
complex dynamic interactions between brain
structures~\citep{SporHone06, BassEtal06, Turk13, DemeEtal19}.

\begin{figure}[tp]
  \centering
  \includegraphics[width=0.65\textwidth]{figs/direction_of_field}
  \caption{\textbf{Neural patterns.} The evolution of fMRI analyses started with
    univariate activation, which refers to
    the average amplitude of BOLD activity evoked by events of an
    experimental condition. Next, multivariate classifiers are trained
    on patterns of activation across voxels to decode distributed
    representations for specific events. The next, resting
    connectivity, is the temporal correlation of one or more seed
    regions with the remainder of the brain during rest. Additionally,
    task-based connectivity examines how these correlations differ by
    cognitive state. Following this increasing trajectory of
    increasing complexity, full connectivity considers all pairwise
    correlations in the brain, most commonly at rest.  Next, dynamic
    full connectivity considers how full connectivity changes over
    time. Continuing this line of reasoning, we expect higher-order network dynamics might provide even richer insights into the neural basis of cognition.}
  \label{fig:direction_of_field}
\end{figure}

@@

To gain insights into the above aspects of the neural code, we
developed a computational framework for estimating dynamic high-order
correlations in timeseries data. This framework provides an important
advance, in that it enables us to examine patterns in higher-order
correlations that are computationally intractable to estimate via
conventional methods.  Given a multivariate timeseries, our framework
provides timepoint-by-timepoint estimates of the first-order
correlations, second-order correlations, and so on (up to tenth-order
correlations in this manuscript).  Our approach combines a
kernel-based method for computing dynamic correlations in timeseries
data with a dimensionality reduction step that projects the resulting
dynamic correlations into a low-dimensional space.  We explored two
dimensionality reduction approaches: principle components
analysis~\citep[PCA;][]{Pear01}, which preserves an approximately invertable
transformation back to the original data; and a second non-invertible
algorithm that explored patterns in eigenvector
centrality~\citep{Land95}.  This latter approach characterizes
correlations between each feature dimension's relative
\textit{position} in the network in favor of the specific activity
histories of different features.

>>>>

\section*{Introduction}
A central goal in cognitive neuroscience is to elucidate the
\textit{neural code}: the mapping between (a) mental states or
cognitive representations and (b) neural activity patterns. One means
of testing models of the neural code is to ask how accurately that
model is able to ``translate'' neural activity patterns into known (or
hypothesized) mental states or cognitive
representations~\citep[e.g.,][]{HaxbEtal01, NormEtal06, TongPrat12,
  MitcEtal08a, KamiTong05, NishEtal11, PereEtal18, HuthEtal12,
  HuthEtal16}.  Training decoding models on different types of neural
features (Fig.~\ref{fig:patterns}a) can also help to elucidate which
specific aspects of neural activity patterns are informative about
cognition-- and, by extension, which types of neural activity patterns
might comprise the neural code.  For example, prior work has used
region of interest analyses to estimate the anatomical locations of
specific neural representations~\citep[e.g.,][]{EtzeEtal09}, or to
compare the relative contributions to the neural code of multivariate
activity patterns versus patterns of dynamic correlations between
neural activity patterns~\citep[e.g.,][]{MannEtal18, FongEtal19}.  An
emerging theme in this literature is that cognition is mediated by
complex dynamic interactions between brain
structures~\citep{SporHone06, BassEtal06, Turk13, DemeEtal19}.

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{figs/patterns}
  \caption{\textbf{Neural patterns.  a. A space of neural analyses}
    Within-brain analyses are carried out within a single brain,
    whereas across-brain analyses compare neural patterns across two
    or more individuals' brains.  Univariate analyses characterize the
    activities of individual units (e.g., nodes, small networks,
    hierarchies of networks, etc.), whereas multivariate analyses
    characterize the patterns of activities across units.  Order 0
    patterns involve individual nodes; order 1 patterns involve
    node-node interactions; order 2 (and higher) patterns relate to
    interactions between homologous networks.  Each of these patterns
    may be static (e.g., averaging over time) or dynamic.
    \textbf{b. Summarizing neural patterns.}  To efficiently compute
    with complex neural patterns, it can be useful to characterize the
    patterns using summary measures.  Dimensionality reduction
    algorithms project the patterns onto lower-dimensional spaces
    whose dimensions reflect weighted combinations of the dimensions
    in the original space.  Graph measures characterize each unit's
    participation in its associated network.}
  \label{fig:patterns}
\end{figure}

@@

To gain insights into the above aspects of the neural code, we
developed a computational framework for estimating dynamic high-order
correlations in timeseries data. This framework provides an important
advance, in that it enables us to examine patterns in higher-order
correlations that are computationally intractable to estimate via
conventional methods.  Given a multivariate timeseries, our framework
provides timepoint-by-timepoint estimates of the first-order
correlations, second-order correlations, and so on (up to tenth-order
correlations in this manuscript).  Our approach combines a
kernel-based method for computing dynamic correlations in timeseries
data with a dimensionality reduction step (Fig.~\ref{fig:patterns}b)
that projects the resulting dynamic correlations into a
low-dimensional space.  We explored two dimensionality reduction
approaches: principle components analysis~\citep[PCA;][]{Pear01},
which preserves an approximately invertable transformation back to the
original data; and a second non-invertible algorithm that explored
patterns in eigenvector centrality~\citep{Land95}.  This latter
approach characterizes correlations between each feature dimension's
relative \textit{position} in the network in favor of the specific
activity histories of different features.
